---
title: "Prediction Assignment Writeup"
author: "Vaishali Katkar"
date: "September 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
To create a report describing how the model is built, how cross validation is use, what is the expected out of sample error is, and why the choices are made. Also prediction model is used to predict 20 different test cases

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Data Loading and Exploratory Analysis

#### Data Set Overview

<B> The training data for this project are available here:</B>

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

<B> The test data are available here: </B>

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har

My special thanks to the above mentioned authors for being so generous in allowing their data to be used for this kind of assignment.

### Course Project Prediction Quiz Portion
Apply machine learning algorithm to the 20 test cases available in the test data above and predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

### Question  

"Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)."

### Input Data
#### Environment Preparation : 
Load R Libraries required : 


```{r}
# Update R Working Directory path : 
setwd("G:/RWrokingDirectory")

# Please make sure that below libraries are installed before running this code.
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```

#### Data Loading and Cleaning :
The next step is loading the dataset from the URL provided above. The training dataset is then partinioned in 2 to create a Training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations. The testing dataset is not changed and will only be used for the quiz results generation.

```{r}
# set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)
```

Both created datasets have 160 variables. Those variables have plenty of NA, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.

```{r}
# Remove the variables which are nearly zero variance.
nearZeroVariance <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -nearZeroVariance]
TestSet <- TestSet[, -nearZeroVariance]
dim(TrainSet)
dim(TestSet)

# Remove the variables which has value mostly as NA
VarNA <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, VarNA==FALSE]
TestSet <- TestSet[, VarNA==FALSE]
dim(TrainSet)
dim(TestSet)

#Remove Identification related columns like - user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp

TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
dim(TestSet)
```

With the cleaning process above, the number of variables for the analysis has been reduced to 54 only

### Correlation Analysis
A correlation among variables is analysed before proceeding to the modeling procedures.

```{r}
CorrelationMatrix <- cor(TrainSet[, -54])
corrplot(CorrelationMatrix, order = "FPC", method = "circle", type = "lower", tl.cex = 0.5, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph above. To make an evem more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step will not be applied for this assignment.

### Prediction Model Building
Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. 
The methods are:- 
<li> Random Forests
<li> Decision Tree
<li> Generalized Boosted Model

as described below :
A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

#### Method Random Forests
```{r}
# model fit
set.seed(12345)

controlRandomForests <- trainControl(method="cv", number=3, verboseIter=FALSE)

modelFitRandForest <- train(classe ~ ., data=TrainSet, method="rf", trControl=controlRandomForests)

modelFitRandForest$finalModel

# prediction on Test dataset
predictRandomForest <- predict(modelFitRandForest, newdata=TestSet)

confMatRandForest <- confusionMatrix(predictRandomForest, TestSet$classe)

confMatRandForest

# plot matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))

```

#### Method: Decision Trees


```{r}
# model fit
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)

# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree

# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))

```

#### Method: Generalized Boosted Model

```{r}
# model fit
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel

# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM

# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))

```

### Applying the Selected Model to the Test Data
The accuracy of the 3 regression modeling methods above are:
a.	Random Forest : 0.9963
b.	Decision Tree : 0.7368
c.	GBM : 0.9839
In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.

```{r}
predictTEST <- predict(modelFitRandForest, newdata=testing)
predictTEST

```